---
title: "Walton_Nate_Exploratory_Data_Analysis"
author: "Nate Walton"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(nycflights13)
library(tidyverse)
```


###5.5 Add new variables with mutate() 

-  mutate()
    - Add new columns that are functions of existing columns
- Easiest way to see all the columns is View().
- If you only want to keep the new variables, use transmute()
- Arithmetic operators: +, -, *, /, ^
    -  If one parameter is shorter than the other, it will be automatically extended to be the same length.
- x / sum(x) calculates the proportion of a total, and y - mean(y) computes the difference from the mean.
-  Logarithms useful to transform for dealing with data that ranges across multiple orders of magnitude.
- lead() and lag() allow you to refer to leading or lagging values
- cumsum(), cumprod(), cummin(), cummax(); and dplyr provides cummean() for cumulative means.
- min_rank(). It does usual type of ranking (e.g. 1st, 2nd, 2nd, 4th)
    - The default gives smallest values the small ranks; use desc(x) to give the largest values the smallest ranks


##5.5.2 Exercises 

1. 


```{r}
1504 %/% 100
1504 %% 100
1504 %/% 100 * 60 + 1504 %% 100
```


flights_times <- mutate(flights,
  dep_time_mins = time2mins(dep_time),
  sched_dep_time_mins = time2mins(sched_dep_time)
)


2. I expect that air_time is the difference between the arrival (arr_time) and departure times (dep_time). In other words, air_time = arr_time - dep_time. There are many flights that air_time != arr_time - dep_time. To fix that I would convert all the times to a date-time to handle overnight flights, and from local time to a common time zone, most likely UTC, to handle flights crossing time-zones.


3. I would expect the departure delay (dep_delay) to be equal to the difference between scheduled departure time (sched_dep_time), and actual departure time (dep_time), dep_time - sched_dep_time = dep_delay.


4.dplyr package has multiple functions for ranking, they handle tied values: row_number(), min_rank(), dense_rank()


5. 
```{r}
c(1 + 1, 2 + 2, 3 + 3, 1 + 4, 2 + 5, 3 + 6, 1 + 7, 2 + 8, 3 + 9, 1 + 10)
```
When adding two vectors recycles the shorter vector’s values to get vectors of the same length.



###5.6 Grouped summaries with summarise()

- summarise(). It collapses a data frame to a single row:
    - summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
    #> # A tibble: 1 x 1
    #>   delay
    #>   <dbl>
    #> 1  12.6
    
- Together group_by() and summarise() provide one of the tools that you’ll use most commonly when working with dplyr: grouped summaries. 
-  explore the relationship between the distance and average delay for each location
    - by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")
- There are three steps to prepare this data:
    - Group flights by destination.
    - Summarise to compute distance, average delay, and number of flights.
    - Filter to remove noisy points and Honolulu airport, which is almost twice as far away as
    the next closest airport.
- tackle the same problem with the pipe, %>%:
    - delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
- Whenever you do any aggregation, it’s always a good idea to include either a count (n()), or a count of non-missing values (sum(!is.na(x)))
- Can switch from %>% to +
-  Lahman package to compute the batting average (number of hits / number of attempts) of every major league baseball player.
- Measures of location: we’ve used mean(x), but median(x)
- Measures of spread: sd(x), IQR(x), mad(x).
- Measures of rank: min(x), quantile(x, 0.25), max(x)
- Measures of position: first(x), nth(x, 2), last(x)
-  To count the number of non-missing values, use sum(!is.na(x)). To count the number of distinct (unique) values, use n_distinct(x).
- Counts and proportions of logical values: sum(x > 10), mean(y == 0)
- If you need to remove grouping, and return to operations on ungrouped data, use ungroup().



##5.6.7 Exercises

1. Being arriving late is more costly to the passenger since it could disrupt the next stages of their travel, such as connecting flights or scheduled meetings. If a departure is delayed without affecting the arrival time, this delay will not have those affects plans nor does it affect the total time spent traveling. If a flight is always 30 minutes late and that delay is known, then it is as if the arrival time is that delayed time.


2. 

not_cancelled %>%
  group_by(tailnum) %>%
  tally(distance)
  

3. If a flight never departs, then it won’t arrive. A flight could also depart and not arrive if it crashes, or if it is redirected and lands in an airport other than its intended destination.


4. The number of cancelled flights increases with the total number of flights per day. The proportion of cancelled flights increases with the average delay of flights.


5. 
```{r}
flights %>%
  group_by(carrier) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(arr_delay))
```

```{r}
filter(airlines, carrier == "F9")
```

Frontier Airlines Inc. Can disentanglingby comparing the average delay of each carrier to the average delay of flights within a route.


6. The sort argument to count() sorts the results in order of n. You could use this anytime you would run count() followed by arrange().



###5.7 Grouped mutates (and filters)

- Grouping is most useful in conjunction with summarise(), but you can also do convenient operations with mutate() and filter()
- A grouped filter is a grouped mutate followed by an ungrouped filter
- You can learn more about useful window functions in the corresponding vignette: vignette("window-functions") 



##5.7.1 Exercises

1. Summary functions (mean()), offset functions (lead(), lag()), ranking functions (min_rank(), row_number()), operate within each group when used with group_by() in mutate() or filter(). Arithmetic operators (+, -), logical operators (<, ==), modular arithmetic operators (%%, %/%), logarithmic functions (log) are not affected by group_by.


2. 

```{r}
flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(n >= 20) %>%
  filter(min_rank(on_time) == 1)
```


3. 
```{r}
flights %>%
  group_by(hour) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)
```

You fly in the morningif you want to avoid delays as much as possible.


4. 

```{r}
flights %>%
  filter(arr_delay > 0) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    arr_delay_prop = arr_delay / arr_delay_total
  ) %>%
  select(
    dest, month, day, dep_time, carrier, flight,
    arr_delay, arr_delay_prop
  ) %>%
  arrange(dest, desc(arr_delay_prop))
```


5. 
```{r}
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))
```


```{r}
lagged_delays %>%
  group_by(origin, dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  facet_wrap(~origin, ncol = 1) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```


6. 
```{r}
standardized_flights <- flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest, origin) %>%
  mutate(
    air_time_mean = mean(air_time),
    air_time_sd = sd(air_time),
    n = n()
  ) %>%
  ungroup() %>%
  mutate(air_time_standard = (air_time - air_time_mean) / (air_time_sd + 1))
```


```{r}
flights %>%
  mutate(mph = distance / (air_time / 60)) %>%
  arrange(desc(mph)) %>%
  select(mph, flight, carrier, flight, month, day, dep_time) %>%
  head(5)
```

```{r}
flights %>%
  mutate(mph = distance / (air_time / 60)) %>%
  arrange(desc(mph)) %>%
  select(
    origin, dest, mph, year, month, day, dep_time, flight, carrier,
    dep_delay, arr_delay
  )
```



```{r}
air_time_delayed <-
  flights %>%
  group_by(origin, dest) %>%
  mutate(
    air_time_min = min(air_time, na.rm = TRUE),
    air_time_delay = air_time - air_time_min,
    air_time_delay_pct = air_time_delay / air_time_min * 100
  )
```


```{r}
air_time_delayed %>%
  arrange(desc(air_time_delay)) %>%
  select(
    air_time_delay, carrier, flight,
    origin, dest, year, month, day, dep_time,
    air_time, air_time_min
  ) %>%
  head() %>%
  print(width = Inf)
```



7. 

```{r}
flights %>%
  # find all airports with > 1 carrier
  group_by(dest) %>%
  mutate(n_carriers = n_distinct(carrier)) %>%
  filter(n_carriers > 1) %>%
  # rank carriers by numer of destinations
  group_by(carrier) %>%
  summarize(n_dest = n_distinct(dest)) %>%
  arrange(desc(n_dest))
```


8. 

```{r}
flights %>%
  # sort in increasing order
  select(tailnum, year, month, day, dep_delay) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(tailnum, year, month, day) %>%
  group_by(tailnum) %>%
  # cumulative number of flights delayed over one hour
  mutate(cumulative_hr_delays = cumsum(dep_delay > 60)) %>%
  # count the number of flights == 0
  summarise(total_flights = sum(cumulative_hr_delays < 1)) %>%
  arrange(total_flights)
```

The exception is flights on the days on which daylight savings started (March 10) or ended (November 3).


























